---
title: "Build-PK-IV - Simple"
description: |
  This tutorial describes a simple pharmacokinetic data building procedure without using additional data processing modules for medications that are intravenously administered.
output:
  distill::distill_article:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE)
knitr::opts_chunk$set(R.options = list(width = 130))
library(EHR)
library(pkdata)
library(lubridate)
library(xaringanExtra)
xaringanExtra::use_clipboard()
```

* See also "Example 1: Quick Data Building with Processed Datasets" in "[2. EHR Vignette for Structured Data](https://cran.r-project.org/web/packages/EHR/vignettes/ehr_vignette_02_str.pdf)" of `EHR` package.

# Introduction
This tutorial describes a simple pharmacokinetic (PK) data building procedure in *EHR2PKPD* for medications that are intravenously (IV) administered. It demonstrates how to build PK data using *Build-PK-IV* when cleaned data for concentration, drug dose, demographic and laboratory datasets are available in the appropriate data format. A comprehensive PK data building procedure with *Build-PK-IV* for IV medications that requires several data processing modules is described in [*Build-PK-IV* - Comprehensive](Build-PK-IV-comprehensive.html) (see Choi *et al.*$^{1}$ for details).

To begin we load the `EHR`. `pkdata`, and  `lubridate` packages. 

```{r load-lib-dir, eval=FALSE}
library(EHR)
library(pkdata)
library(lubridate)
```

# Quick Data Building with Processed Datasets

There are **three basic steps** to build a PK dataset.

## (1) Define directories
* a directory for the raw data (`rawDataDir` in the example below)
* a directory for interactive checking output files (`checkDir` in the example below)

```{r define-directories}
rawDataDir <- system.file("examples", "str_ex1", package="EHR")
td <- tempdir()
checkDir <- file.path(td, 'check1')
dir.create(checkDir)
```

## (2) Load cleaned and appropriately formatted data files

Four types of files are used in the *Build-PK-IV* module:

* an IV dosing file
* a drug concentration file
* a demographic file
* a laboratory file (optional)

```{r simp-in}
ivdose.data <- read.csv(file.path(rawDataDir,"IVDose_DATA_simple.csv"),stringsAsFactors = FALSE)
head(ivdose.data, 3)
conc.data <- read.csv(file.path(rawDataDir,"Concentration_DATA_simple.csv"),stringsAsFactors = FALSE)
head(conc.data, 3)
demo <- read.csv(file.path(rawDataDir,"Demographics_DATA_simple.csv"),stringsAsFactors = FALSE)
head(demo, 3)
creat.data <- read.csv(file.path(rawDataDir,"Creatinine_DATA_simple.csv"),stringsAsFactors = FALSE)
head(creat.data, 3)
```

All of the datasets are in the appropriate data format and include the variable `patient_id`, a unique patient-level ID. The concentration and demographic files also contain a `patient_visit_id` variable, which is a unique visit-level ID. <!--In order to run the *Build-PK-IV* module, these variables must be renamed, as we demonstrate in the next section.

## (3) Rename ID variables with standardized names
The `EHR` package modules use a standardized naming convention for patient identification (ID) variables. We rename the unique patient-level ID from `patient_id` to `mod_id` and the visit-level ID from `patient_visit_id` to `mod_id_visit`. If there is only a single visit/course per subject, the unique patient-level ID and visit-level ID can be the same, however both `mod_id` and `mod_id_visit` should be defined.

```{r simp-rename}
# rename ID variables
#names(conc.data)[1:2] <- names(demo)[1:2] <- c("mod_id", "mod_id_visit")
#names(creat.data)[1] <- names(ivdose.data)[1] <- "mod_id"
```
-->

## (3) Build a final PK dataset with the function `run_Build_PK_IV()`

The following arguments are used in the `run_Build_PK_IV` function:

* `conc`: drug concentration data
* `conc.columns`: a named list that should specify columns in concentration data
* `dose`: IV dose data
* `dose.columns`:  a named list that should specify columns in dose data
* `demo.list`: demographic data; if provided, ‘id’ is required in `demo.columns`
* `demo.columns`: a named list that should specify columns in demographic data
* `lab.list`: laboratory data
* `lab.columns`: a named list that should specify columns in lab data
* `check.path`: (optional) file path where the generated files for data checking are stored, and the corresponding data files with fixed data exist

In this tutorial, we describe only arguments relevant to this example. A detailed description of all arguments can be found in the `EHR` package documentation for `run_Build_PK_IV()`.

```{r sim-build-pk-iv, collapse = TRUE}
simple_pk_dat <- run_Build_PK_IV(
    conc=conc.data,
    conc.columns = list(id = 'patient_id', datetime = 'date.time', druglevel = 'conc.level', 
                        idvisit = 'patient_visit_id'),
    dose=ivdose.data,
    dose.columns = list(id = 'patient_id', date = 'date.dose', infuseDatetime = 'infuse.time', 
                        infuseDose = 'infuse.dose', infuseTimeExact= 'infuse.time.real', 
                        bolusDatetime = 'bolus.time', bolusDose = 'bolus.dose', 
                        gap = 'maxint', weight = 'weight'),
    demo.list = demo,
    demo.columns = list(id = 'patient_id', idvisit = 'patient_visit_id'),
    lab.list = list(creat.data),
    lab.columns = list(id = 'patient_id', datetime = 'date.time'),
    check.path=checkDir)

```

<!--
```{r}
# simple_pk_dat <- run_Build_PK_IV(
#     conc=conc.data,
#     conc.columns = list(id = 'mod_id', datetime = 'date.time', druglevel = 'conc.level', 
#                         idvisit = 'mod_id_visit'),
#     dose=ivdose.data,
#     dose.columns = list(id = 'mod_id', date = 'date.dose', infuseDatetime = 'infuse.time', 
#                         infuseDose = 'infuse.dose', infuseTimeExact= 'infuse.time.real', 
#                         bolusDatetime = 'bolus.time', bolusDose = 'bolus.dose', 
#                         gap = 'maxint', weight = 'weight'),
#     demo.list = demo,
#     demo.columns = list(id = 'mod_id', idvisit = 'mod_id_visit'),
#     lab.list = list(creat.data),
#     lab.columns = list(id = 'mod_id', datetime = 'date.time'),
#     check.path=checkDir)
```
-->

The `run_Build_PK_IV()` function generates an automatic message that provides information about the data processing and the final dataset including the variables, the sample size, and missingness.

Below we show the final PK dataset.

```{r sim-build-pk-iv-out}
head(simple_pk_dat,15)
```

This dataset includes the `patient_visit_id` variable and standard NONMEM formatted variables (for details of data items, see [PK-Data-IV-Dosing](PK-Data-IV-Dosing.html)).

* time - time of dosing or concentration event
* amt - dose amount administered (NA for concentration records)
* dv - dependent variable; i.e., observed concentration (NA for dosing records)
* rate - rate of drug administration (e.g., rate=0 for bolus doses)
* mdv - missing dependent variable (dv) indicator (e.g., 0 = not missing dv, 1 = missing dv)
* evid - event ID (e.g., 0 = observation, 1 = dose event)

If demographic data is provided, the demographic variables will also be included. 

# Example with only two datasets

The `run_Build_PK_IV()` can also be used with only two datasets:

* an IV dosing file
* a file with drug concentration, laboratory values, demographics, etc.

To illustrate this, we generate an example dataset with concentration, laboratory values, and demographics combined, which is shown below.

```{r merge and sim-conc-comb}
mrg0 <- merge(conc.data,creat.data,by=c('patient_id','date.time'),all=TRUE)
mrg1 <- merge(mrg0,demo,by=c('patient_id','patient_visit_id'),all=TRUE)
conc.combined <- mrg1[!is.na(mrg1$conc.level),]

head(conc.combined,3)
```

The following arguments are used in the run_Build_PK_IV function:

* `conc`: drug concentration along with other data (laboratory, demographic, etc.)  
* `conc.columns`: a named list that should specify columns in concentration data  
* `dose`: IV dose data  
* `dose.columns`: a named list that should specify columns in dose data
* `pk.vars`: variables to include in the returned PK data. If `pk.vars` includes ‘date’, the output generates its original date-time to which the ‘time’ is mapped. Users can use `pk.vars` to include variables for demographics or laboratory values that are already merged with the concentration dataset. 

```{r, collapse = TRUE}
simple_pk_dat2 <- run_Build_PK_IV(
    conc = conc.combined,
    conc.columns = list(id='patient_id', datetime='date.time', druglevel='conc.level'),
    dose = ivdose.data,
    dose.columns = list(id='patient_id', date = 'date.dose', infuseDatetime = 'infuse.time',
                        infuseDose = 'infuse.dose', infuseTimeExact ='infuse.time.real',
                        bolusDatetime = 'bolus.time', bolusDose = 'bolus.dose',
                        gap = 'maxint', weight = 'weight'),
    pk.vars = c('date','weight','height','ageatsurgery','creat',
                'stat_sts','cpb_sts','length_of_icu_stay'))
```

```{r}
# the final PK dataset
head(simple_pk_dat2,15)
```

# References
1. Choi L, Beck C, McNeer E, Weeks HL, Williams ML, James NT, Niu X, Abou-Khalil BW, Birdwell KA, Roden DM, Stein CM. Development of a System for Post-marketing Population Pharmacokinetic and Pharmacodynamic Studies using Real-World Data from Electronic Health Records. Clinical Pharmacology & Therapeutics. 2020 Apr;107(4):934-43. doi: 10.1002/cpt.1787.
