---
title: "PK Data for IV Dosing"
description: |
  This tutorial introduces the population pharmacokinetic modeling novice to the data format appropriate for most PK model fitting software systems with a focus on intravenous dosing data.
author:
  - name: Nathan T. James
output:
  distill::distill_article:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(R.options = list(width = 120))
library(EHR)
library(pkdata)
library(lubridate)
```

**Under construction.**

# Introduction
We briefly describe a data format typically used in pharmacokinetics (PK) and pharmacodynamics (PD) modeling, especially relevant to our example intravenous (IV) dosing data. We describe the data form using data items required in NONMEM, which is acceptable in most PK software. More extensive description of data format appropriate for PK software can be found in the manual of each software such as the MONMEM manual$^{1}$.

# Population PK/PD Data Format

Here is an example of data for intravenously administered drugs in a suitable format for PK analysis using most PK software including NONMEM.


EXAMPE DATA HERE

```{r echo=TRUE}
library(EHR)
# Data generating function for examples
mkdat <- function() {
  npat=3
  visits <- floor(runif(npat, min=2, max=6))
  id <- rep(1:npat, visits)
  dt <- as.POSIXct(paste(as.Date(sort(sample(700, sum(visits))), 
                                 origin = '2019-01-01'), '10:00:00'), tz = 'UTC') 
  + rnorm(sum(visits), 0, 1*60*60)
  dose_morn <- sample(c(2.5,5,7.5,10), sum(visits), replace = TRUE)
  conc <- round(rnorm(sum(visits), 1.5*dose_morn, 1),1)
  ld <- dt - sample(10:16, sum(visits), replace = TRUE) * 3600
  ld[rnorm(sum(visits)) < .3] <- NA
  age <- rep(sample(40:75, npat), visits)
  weight <- rep(round(rnorm(npat, 180, 20)),visits)
  hgb <- round(rep(rnorm(npat, 10, 2), visits),1)
  data.frame(id, dt, dose_morn, conc, age, weight, hgb, ld)
}

# Make example data
set.seed(20)
dat <- mkdat()
ex <- dat
dat2 <- dat[,-8]

library(EHR)
# Build PK data without last-dose times
exampledata <- run_Build_PK_Oral(x = dat2,
                  idCol = "id",
                  dtCol = "dt",
                  doseCol = "dose_morn",
                  concCol = "conc",
                  ldCol = NULL,
                  first_interval_hours = 336,
                  imputeClosest = NULL)
exampledata <- exampledata[,c("id","time","amt","dv","mdv","evid","addl","II","age")]
```

```{r echo=TRUE}
exampledata
```

This simple example has two types of data rows, dose events (information concerning dosing) and concentration events (information concerning blood concentration levels for the drug of interest).  The columns indicate the following information:

1. `id`: unique identifier for each individual in the dataset.
1. `time`: time of either dosing or concentration measurement.
1. `amt`: dose amount; `NA` for concentration events.
1. `dv`: drug blood concentration value, which is DV (dependent variable) as NONMEM data item; `NA` for a dose event.
1. `mdv`: missing dependent variable; 1 for indicating that there is no dependent variable (in this case, blood concentration), 0 for dependent variable.
1. `evid`: event ID; 1 for indicating dose event (amt, II, and addl for this record will be used for the drug dose information if evid = 1), 0 for observation (or dependent variable if mdv = 0).
1. `weight`: an example covariate; each column can have covariates - these can be time varying or fixed within individual.

DESCRIPTION OF EXAMPE HERE

Taking the first row as an example, which is a dose event, a dose of 10 mg is taken at time 0 then 27 additional doses (the same 10 mg) are taken every 12 hours (time can take any unit, or the same unit of time variable).  If we calculate each of these 27 doses then we see that the final dose is given at `time = 324`, 12 hours before the measured concentration of 16.1 in row 2.  Of course this same dosing information could be expressed in 28 rows, (10 mg at time 0, 10 mg at time 12, 10 mg at time 24, etc.) but this more compact form is preferable in the presence of a regular dosing interval.

This PK data form will be generated using *Build-PK-IV* module (see [*Build-PK-IV* : Simple](Build-PK-IV-simple.html) or [*Build-PK-IV* : Comprehensive](Build-PK-IV-comprehensive.html) ). 

# References  
1. Beal, S.L., Sheiner, L.B., Boeckmann, A.J. & Bauer, R.J. (Eds). NONMEM 7.5 users guides (1989–2020).
