---
title: "PK Data for IV Dosing"
description: |
  This tutorial introduces the population pharmacokinetic modeling novice to the data format appropriate for most PK model fitting software systems with a focus on IV dosing data.
author:
  - name: Nathan T. James
output:
  distill::distill_article:
    toc: true
    toc_depth: 2
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(R.options = list(width = 120))
library(EHR)
library(pkdata)
library(lubridate)
```

* See also "[2. EHR Vignette for Structured Data](https://cran.r-project.org/web/packages/EHR/vignettes/ehr_vignette_02_str.html)" in `EHR` package.

# Introduction
We briefly describe a data format typically used in pharmacokinetics (PK), pharmacodynamics (PD) modeling.  There is a brief example dataset and a breakdown of the information contained in each column and their appropriate formatting.

# Population PK/PD Data Format for IV Dosing

Here is an example of PK data for orally administered drugs in a suitable format for analysis in NONMEM.

```{r echo=TRUE}
library(EHR)
# Data generating function for examples


```

Under construction.

# References  
1. NONMEM manual
