---
title: "Pro-Med-Str - Part I"
description: |
  This tutorial describes how to process structured medication data, especially focusing on intravenously given dose data.
output:
  distill::distill_article:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(R.options = list(width = 130))
library(EHR)
```

* See also "Pro-Med-Str Part I: IV dose data" in "[2. EHR Vignette for Structured Data](https://cran.r-project.org/web/packages/EHR/vignettes/ehr_vignette_02_str.html)" of `EHR` package.

# Introduction

This tutorial describes how to use the *Pro-Med-Str* - Part I module in *EHRtoPKPD* system to process intravenous (IV) dose data (see Choi *et al.*$^{1}$ for details).

To begin we load the `EHR` package,

```{r load-lib-dir, eval=FALSE}
# load EHR package
library(EHR)
```

# Input Raw Dose Data

We will use example dose data to demonstrate the *Pro-Med-Str Part I* module. The raw data is shown below.

```{r}
mar.in0 <- read.csv(system.file("examples", "str_ex2","MAR_DATA.csv", package="EHR"), check.names = FALSE)
head(mar.in0)
```

This data consists of a patient ID, date, time, medication name, dosage, route, frequency, and medication given.

The patient ID may need to be renamed so that all input datasets have the same name for the patient ID. This is necessary when combining the datasets to create a crosswalk between the original ID variables and the new ID variables used in the *Pro-Med-Str Part I* module. We demonstrate how to rename the patient ID variable below.


```{r}
mar.new <- dataTransformation(mar.in0, rename = c('Uniq.Id' = 'subject_uid'))
```

# Preparing Raw Dose Data

In practice, the dose data will need to be combined with other input datasets. This process involves creating a crosswalk between original ID variables and new ID variables. The new ID variables that are required to be the same across all datasets are `mod_id`, `mod_visit`, and `mod_id_visit`. See "[2. EHR Vignette for Structured Data](https://cran.r-project.org/web/packages/EHR/vignettes/ehr_vignette_02_str.html)" of `EHR` package and "[Build-PK-IV - Comprehensive Workshop](https://choileena.github.io/Build-PK-IV-comprehensive.html)" for examples of this process.

For simplicity, we will skip this step in this tutorial.

We need to save our dataset as an RDS file using `saveRDS` as shown below. Here, we create a temporary directory to store the files using `tempdir`. We define two directories, `dataDir` for processed data and `checkDir` containing files used for interactive checking (optional). Instead of using a temporary directory, `dataDir` and `checkDir` can be specific directories on your computer.

```{r}
# define 2 directories
td <- tempdir()
dataDir <- file.path(td, 'data') # directory for processed data
dir.create(dataDir)
checkDir <- file.path(td, 'checks') # directory for interactive checking
dir.create(checkDir)
saveRDS(mar.new, file=file.path(dataDir,"mar_new.rds"))
```

The next section demonstrates how to use the `run_MedStrI` function to run the *Pro-Med-Str* module.

# Running `run_MedStrI()`

* IV dose data can be in various forms, which may need to be pre-processed. IV dose data can be obtained from different data sources when using electronic health records (EHRs), but we illustrate this module using an example dataset (called "MAR") that comes from one data source for simplicity (see "[Build-PK-IV - Comprehensive Workshop](https://choileena.github.io/Build-PK-IV-comprehensive.html)" for a case when dose data are obtained from two data sources).

* `run_MedStrI()` is the main function to process IV dose data in *Pro-Med-Str* Part I module.
* The module can be semi-interactive -- it generates several files to check potential data errors and get feedback from an investigator. If corrected information ('fix' files) are provided, the module should be re-run to incorporate the corrections. 

* `run_MedStrI()` can take two types of raw IV dose data (e.g., flow data and MAR data). In this tutorial, we describe arguments only relevant to processing one of these datasets. A detailed description of all arguments can be found in the `EHR` package manual of `run_MedStrI()`.
    + `mar.path`: file path where the MAR data exist
    + `mar.columns`: a named list that should specify columns in MAR data; ‘id’, ‘datetime’ and ‘dose’ are required. ‘drug’, ‘weight’, ‘given’ may also be specified. ‘datetime’ is date and time for data measurement, which can refer to a single date-time variable (datetime = 'date_time') or two variables holding date and time separately (e.g., datetime = c('Date', 'Time')). ‘dose’ can also be given as a single variable or two variables. If given as a single column, the column's values should contain dose and units such as '25 mcg'. If given as two column names, the dose column should come before the unit column (e.g., dose = c('doseamt', 'unit')). If ‘drug’ is present, the ‘medchk.path’ argument should also be provided, which contain a csv file with a list of drug names that should be used to subset the relevant drugs. The ‘given’ variable should be used in conjunction with the ‘medGivenReq’ argument.
    + `medGivenReq`: values in the ‘given’ column should equal “Given”; if this is FALSE, NA values are also allowed
    + `medchk.path`: file path with list of medication names and variants in MAR data to be used 
    + `check.path`: file path where the generated files for data checking are stored, and the corresponding data files with fixed data exist
    + `failunit_fn`: filename stub for records with units other than those specified with `infusion.unit` and `bolus.unit`
    + `infusion.unit`: string specifying units for infusion doses to check correct unit (default: 'mcg/kg/hr')
    + `bolus.unit`: string specifying units for bolus doses to check correct unit (default: 'mcg')
    + `bol.rate.thresh`: upper bound for retaining bolus doses. Bolus units with a rate above the threshold are dropped (default: Inf; i.e., keep all bolus doses)
    + `rateunit`: string specifying units for infusion rate (default: 'mcg/hr')
    + `ratewgtunit`: string specifying units for infusion rate per body weight (default: 'mcg/kg/hr')
    + `weightunit`: string specifying units for subject body weight (default: 'kg')
    + `drugname`: drug name of interest (e.g., dex, fent)

Below we show how we would run `run_MedStrI()` using the cleaned IV dose dataset, "mar_new.rds", as input.

```{r Pro-Med-Str1}
# define parameters
drugname <- 'fent'

ivdose.out <- run_MedStrI(
    mar.path=file.path(dataDir,"mar_new.rds"),
    mar.columns = list(id='subject_uid', datetime=c('Date','Time'), dose='med:dosage', drug='med:mDrug', given='med:given'),
    medGivenReq = TRUE,
    medchk.path = file.path(system.file("examples", "str_ex2", package="EHR"), sprintf('medChecked-%s.csv', drugname)),
    check.path=checkDir, 
    failunit_fn = 'Unit',
    infusion.unit = 'mcg/kg/hr',
    bolus.unit = 'mcg',
    bol.rate.thresh = Inf,
    rateunit = "mcg/hr",
    ratewgtunit = "mcg/kg/hr",
    weightunit = "kg",
    drugname = drugname
    )
```

After `run_MedStrI()` is executed, it provides messages such as those shown above, which contain some information about data checking.

# Output of `run_MedStrI()`

```{r}
head(ivdose.out)
```

* The output from the *Pro-Med-Str* - Part I module is a `data.frame` with the following columns:  

  - `subject_uid`: the new id that will be used to merge datasets.
  - `date.dose`: dose given date.
  - `infuse.time.real`: infusion dose time recorded in the raw data.
  - `infuse.time`: infusion dose time processed based on continuous infusion time; this time is used in the PK data building).  
  - `infuse.dose`: infusion dose.
  - `bolus.time`: bolus dose time.
  - `bolus.dose`: bolus dose.
  - `given.dose`: variable in the raw data stating the dose was given.
  - `maxint`: infusion recording interval (e.g., 15 min, 1 hour).
  - `weight`: subject body weight recorded near infusion time.

This output is the input of `run_Build_PK_IV()` function in *Build-PK-IV* module. In the above output, all variables except `maxint` and `weight` are the necessary variables that should be reserved to use `run_Build_PK_IV()`.

# References
1. Choi L, Beck C, McNeer E, Weeks HL, Williams ML, James NT, Niu X, Abou-Khalil BW, Birdwell KA, Roden DM, Stein CM. Development of a System for Post-marketing Population Pharmacokinetic and Pharmacodynamic Studies using Real-World Data from Electronic Health Records. Clinical Pharmacology & Therapeutics. 2020 Apr;107(4):934-43. doi: 10.1002/cpt.1787.

