---
title: "Pro-Drug Level"
description: |
  This tutorial describes how to process drug concentration data using *Pro-Drug Level* module in the system.
author:
  - name: Michael L. Williams
output:
  distill::distill_article:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(R.options = list(width = 120))
library(EHR)
```

* See also "Pro-Drug Level" in "[2. EHR Vignette for Structured Data](https://cran.r-project.org/web/packages/EHR/vignettes/ehr_vignette_02_str.html)" of `EHR` package.

# Introduction

This tutorial describes how to use the *Pro-Drug Level* module to process drug concentration data (see Choi *et al.*$^{1}$ for details).

To begin we load the `EHR` package and create some directories to work from. 

```{r load-lib-dir, eval=TRUE}
# load EHR package
library(EHR)

td <- tempdir()
dir.create(file.path(td, 'checks'))
checkDir <- file.path(td, 'checks') # directory for interactive checking
dir.create(file.path(td, 'data'))
dataDir <- file.path(td, 'data') # directory for processed data
rawDataDir <- system.file("examples", "str_ex2", package="EHR")
drugname <- 'fent'
LLOQ <- 0.05
```



# Drug Level Data

Drug level data is an essential part of pharmacokinetic (pk) data.  It comprises the dependent variable in the model, the response elicited by medication dosing.  It is generally measured as drug concentration in the blood calculated in mass per volume.

```{r}
# helper function used to make subject_id
sampId <- function(x) {
  # remove leading zeroes or trailing periods
  subid <- gsub('(^0*|\\.$)', '', x)
  # change _ to .
  gsub('_([0-9]+[_].*)$', '.\\1', subid)
}

# read SampleConcentration_DATA.csv
conc.in <- readTransform(file.path(rawDataDir, "SampleConcentration_DATA.csv"),
  modify = list(
    subid = expression(sampId(name)),
    subject_id = expression(as.numeric(sub('[_].*', '', subid))),
    samp = expression(sub('[^_]*[_]', '', subid)),
    name = NULL,
    data_file = NULL,
    subid = NULL
    )
  )
head(conc.in)
```

Each row in the data represents an observation of fentanyl blood level in an individual.  Each is associated with a sample number, the details of which are in another file.

```{r}
# read SampleTimes_DATA.csv
samp.in <- readTransform(file.path(rawDataDir, "SampleTimes_DATA.csv"),
    rename = c('Study.ID' = 'subject_id'),
    modify = list(samp = expression(as.numeric(sub('Sample ', '', Event.Name)))))
head(samp.in)
```

The goal of *Pro-Druglevel* is to make this information suitable for merging medication dosing data to make it suitable for population pharmacokinetic analysis. As such, it is recommended that users develop an understanding of *Pro-Med-NLP* and as the *Pro-Druglevel* module functions in concert with that modules.

# Additional Data

Load drug dosing data to demonstrate how the data is merged.

```{r mar-in}
mar.in0 <- read.csv(file.path(rawDataDir, "MAR_DATA.csv"), check.names = FALSE)
mar.in <- dataTransformation(mar.in0, rename = c('Uniq.Id' = 'subject_uid'))
head(mar.in)
```

# Establishing a unified ID system

We create a new ID system to aid de-identification of the dataset before further analysis.

```{r}
demo.in <- readTransform(file.path(rawDataDir, "Demographics_DATA.csv"))



data <-  list(demo.in,
              mar.in,
              conc.in,
              samp.in)

# define list of vectors or character strings that identify the ID variables
idcols <-  list(c('subject_id','subject_uid'),
                'subject_uid',
                'subject_id',
                'subject_id')

# merge all IDs from cleaned datasets and create new ID variables
id.xwalk <- idCrosswalk(data, idcols, visit.id="subject_id", uniq.id="subject_uid")
saveRDS(id.xwalk, file=file.path(dataDir,"module_id_xwalk.rds"))


demo.cln <- pullFakeId(demo.in, id.xwalk)
head(demo.cln)
saveRDS(demo.cln, file=file.path(dataDir,"demo.rds"))

mar.cln <- pullFakeId(mar.in, id.xwalk, firstCols = 'mod_id', uniq.id = 'subject_uid')
saveRDS(mar.cln, file=file.path(dataDir,"mar.rds"))

conc.cln <- pullFakeId(conc.in, id.xwalk,
    firstCols = c('record_id', 'mod_id', 'mod_visit', 'mod_id_visit', 'samp'),
    orderBy = 'record_id',
    uniq.id = 'subject_uid')
saveRDS(conc.cln, file=file.path(dataDir,"conc.rds"))

samp.cln <- pullFakeId(samp.in, id.xwalk,
    firstCols = c('mod_id', 'mod_visit', 'mod_id_visit', 'samp'), 
    orderBy = c('mod_id_visit','samp'),
    uniq.id = 'subject_uid')
saveRDS(samp.cln, file=file.path(dataDir,"samp_mod_id.rds"))
```

# Processing the Datasets

The three datasets are now ready for processing, `run_Druglevel` will configure the drug level data by merging concentration and sample data.  A detailed descriptions for `run_MedStrI` can be found in the module *Pro-Med-Str*.

```{r}
conc.out <- run_DrugLevel(conc.path=file.path(dataDir,"conc.rds"),
    conc.select=c('mod_id','mod_id_visit','samp','fentanyl_calc_conc'),
    conc.rename=c(fentanyl_calc_conc = 'conc.level', samp= 'event'),
    conc.mod.list=list(mod_id_event = expression(paste(mod_id_visit, event, sep = '_'))),
    samp.path=file.path(dataDir,"samp_mod_id.rds"),
    samp.mod.list=list(mod_id_event = expression(paste(mod_id_visit, samp, sep = '_'))),
    check.path=checkDir,
    drugname=drugname,
    LLOQ=LLOQ)
head(conc.out)
```

```{r}
ivdose.out <- run_MedStrI(mar.path=file.path(dataDir,"mar.rds"),
                          medchk.path=file.path(rawDataDir, "medChecked-fent.csv"),
                          check.path=checkDir,
                          drugname=drugname)
```

# Building the final dataset

Pass the two processed datasets to `run_Build_PK_IV` now.  The drug level data will be appropriately represented in the final dataset

```{r}
pk_dat <- run_Build_PK_IV(
    conc=conc.out,
    dose=ivdose.out,
    pk.vars=c('conc','mod_id_visit', 'time', 'dose', 'rate', 'event',
              'other', 'multiple.record', 'date', 'mod_id'),
    drugname=drugname,
    check.path=checkDir)

head(pk_dat)
```

This data is now of a form suitable for population pharmacokinetic analysis.

# References
1. Choi L, Beck C, McNeer E, Weeks HL, Williams ML, James NT, Niu X, Abou-Khalil BW, Birdwell KA, Roden DM, Stein CM. Development of a System for Post-marketing Population Pharmacokinetic and Pharmacodynamic Studies using Real-World Data from Electronic Health Records. Clinical Pharmacology & Therapeutics. 2020 Apr;107(4):934-43. doi: 10.1002/cpt.1787.

