---
title: "Pro-Med-Str - Part II"
description: |
  This lecture describes how to process structured medication data, especially focusing on e-prescription data.
author:
  - name: Elizabeth McNeer
output:
  distill::distill_article:
    toc: true
    toc_depth: 1
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(EHR)
```

Here add links to vignettes

# Introduction

This lecture describes how to use the Pro-Med-Str - Part II module to process structured e-prescription data in *EHRtoPKPD*.

The major tasks the module performs are as follows:

* Creating numeric variables for strength, dose, and frequency
* Calculating daily dose
* Removing duplicate daily doses for a patient

# Raw E-Prescription Data

To use this module, all prescriptions must be for only one drug. Different names, such as brand names and generic names, for the same drug are allowed (e.g., Lamictal and lamotrigine). The data used in this module must include columns for ID, date, strength, dose amount, and frequency. If a description column is included, this module will attempt to extract the strength from the description column in cases where the strength is missing.

Below is example e-prescription data from the `EHR` package including columns for ID, drug name, dose, frequency, date, strength, and description.

```{r eRX-dat}
rawDataDir <- system.file("examples", "str_ex2", package="EHR")
(eRX <- read.csv(file.path(rawDataDir,"e-rx_DATA.csv"),stringsAsFactors = FALSE))
```

To get a daily dose for each patient, we multiply strength\*dose\*frequency. In order to do this, the STRENGTH_AMOUNT, RX_DOSE, and FREQUENCY variables need to be converted to numeric. Strengths that include units will have the units removed (e.g., numeric strength for ID4 will be 200), frequencies will be converted to the equivalent number of times per day that the medication is taken (e.g., bid = 2, twice a day = 2, qam = 1), and doses will have words like "tabs" removed (e.g., numeric dose for ID5 will be 2). For ID6, separate doses are written for breakfast, lunch, and dinner, so the numeric dose will be 4 (1.5+1+1.5), and the daily dose will be calculated as strength\*dose. ID3 is missing a value for STRENGTH_AMOUNT, but we can use the strength that is present in the DESCRIPTION column. In the next section, we show how the `run_MedStrII` function in the `EHR` package takes care of all of these tasks for us and calculates a daily dose.

# Processing the Data

We begin by loading the `EHR` package.

```{r,eval=FALSE}
library(EHR)
```

The e-prescription data can be processed by the `run_MedStrII` function using: 

```{r Pro-Med-Str2}
eRX.out <- run_MedStrII(file.path(rawDataDir,"e-rx_DATA.csv"),
    select = c('GRID','MED_NAME','RX_DOSE','FREQUENCY','ENTRY_DATE','STRENGTH_AMOUNT','DESCRIPTION'),
    rename = c('ID','MED_NAME','RX_DOSE','FREQUENCY','ENTRY_DATE','STRENGTH_AMOUNT','DESCRIPTION'))
```

The following arguments are used in the `run_MedStrII` function:

* `file`: file name of prescription data
* `select`: the names of the columns to select
* `rename`: new column names; the default are the names required for the underlying functions, `processErx` and `processErxAddl`

```{r}
eRX.out
```

In the above example, daily dose was calculated for the first 5 patients by multiplying strength\*dose\*freq.num, and a redundant daily dose was removed for the patient with ID2. In order to calculate a daily dose for the patient with ID3, the strength of 100 from the description was used because STRENGTH_AMOUNT was missing. For the patient with ID6, the dose amounts of 1.5, 1, and 1.5 are added together to get a dose of 4, and the daily dose is calculated as strength\*dose.
