---
title: "Pro-Demographic"
description: |
  This tutorial describes how to process demographic data using *Pro-Demographic* module in the system.
author:
  - name: Michael L. Williams
output:
  distill::distill_article:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(R.options = list(width = 120))
library(EHR)
```

* See also "Pro-Demographic" in "[2. EHR Vignette for Structured Data](https://cran.r-project.org/web/packages/EHR/vignettes/ehr_vignette_02_str.html)" of `EHR` package.

# Introduction

This tutorial describes how to use the *Pro-Demographic* module to process demographic data (see Choi *et al.*$^{1}$ for details).

To begin we load the `EHR` package. 

```{r load-lib-dir, eval=FALSE}
# load EHR package and dependency
library(EHR)
library(medExtractR)
```




# References
1. Choi L, Beck C, McNeer E, Weeks HL, Williams ML, James NT, Niu X, Abou-Khalil BW, Birdwell KA, Roden DM, Stein CM. Development of a System for Post-marketing Population Pharmacokinetic and Pharmacodynamic Studies using Real-World Data from Electronic Health Records. Clinical Pharmacology & Therapeutics. 2020 Apr;107(4):934-43. doi: 10.1002/cpt.1787.

