---
title: "Pro-Laboratory"
description: |
  This tutorial describes how to process laboratory data using *Pro-Laboratory* module in the system.
author:
  - name: Elizabeth McNeer
output:
  distill::distill_article:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(R.options = list(width = 120))
library(EHR)
```

* See also "Pro-Laboratory" in "[2. EHR Vignette for Structured Data](https://cran.r-project.org/web/packages/EHR/vignettes/ehr_vignette_02_str.html)" of `EHR` package.

# Introduction

This tutorial describes how to use the *Pro-Laboratory* module to process laboratory data (see Choi *et al.*$^{1}$ for details).

To begin we load the `EHR` package. 

```{r load-lib-dir, eval=FALSE}
# load EHR package
library(EHR)
```

# Laboratory Data

We will use example creatinine lab data to demonstrate the *Pro-Laboratory* module. The raw data is shown below.

```{r}
creat.in <- read.csv(system.file("examples", "str_ex2","Creatinine_DATA.csv", package="EHR"))
head(creat.in)
```

This data consists of a patient ID, date, time, and the creatinine level.

The patient ID may need to be renamed so that all input datasets have the same name for the patient ID. This is necessary when combining the datasets to create a crosswalk between the original ID variables and the new ID variables used in the *Pro-Laboratory* module. See "[2. EHR Vignette for Structured Data](https://cran.r-project.org/web/packages/EHR/vignettes/ehr_vignette_02_str.html)" of the `EHR` package for more information. We demonstrate how to rename the patient ID variable below.

```{r}
creat.new <- dataTransformation(creat.in, rename = c('Subject.uniq' = 'subject_uid'))
```

# Preparing Raw Laboratory Data

In practice, the laboratory data will need to be combined with other input datasets. This process involves creating a crosswalk between original ID variables and new ID variables. The new ID variables that are required to be the same across all datasets are `mod_id`, `mod_visit`, and `mod_id_visit`. See "[2. EHR Vignette for Structured Data](https://cran.r-project.org/web/packages/EHR/vignettes/ehr_vignette_02_str.html)" of `EHR` package and "[Build-PK-IV - Comprehensive Workshop](https://choileena.github.io/Build-PK-IV-comprehensive.html)" for examples of this process.

For simplicity, we will skip this step in this tutorial.

We need to save our dataset as an RDS file using `saveRDS` as shown below. Here, we create a temporary directory to store the file using `tempdir`; however, `dataDir` can be a specific directory on your computer.

```{r}
td <- tempdir()
dir.create(file.path(td, 'data2'))
dataDir <- file.path(td, 'data2')
saveRDS(creat.new, file=file.path(dataDir,"creat_new.rds"))
```

The next section demonstrates how to use the `run_Labs` function to run the *Pro-Laboratory* module.

# Running `run_Labs`

The following arguments must be specified:

- `lab.path`: The file path where the laboratory data exist. It must be an RDS file.
- `lab.select`: The list of variables in the laboratory data to be retained.
- `lab.mod.list`: A list containing modifications to variables in the laboratory data.

Below we show how we would run `run_Labs` using the example laboratory data from above.

```{r}
creat.out <- run_Labs(lab.path=file.path(dataDir,"creat_new.rds"),
    lab.select = c('subject_uid','date.time','creat'),
    lab.mod.list = list(date.time = expression(parse_dates(fixDates(paste(date, time))))))
```

In the above code, the `lab.mod.list` argument specifies a modification to our dataset to include a `date.time` variable, which is created by combining the original `date` and `time` variables. The `lab.select` argument says that we want to keep the `subject_uid`, `date.time`, and `creat` variables.

# Output of `run_Labs`

```{r}
head(creat.out)
```

This data can be merged with data from other modules. See the "[Build-PK-IV - Comprehensive Workshop](https://choileena.github.io/Build-PK-IV-comprehensive.html)" for an example.

# References
1. Choi L, Beck C, McNeer E, Weeks HL, Williams ML, James NT, Niu X, Abou-Khalil BW, Birdwell KA, Roden DM, Stein CM. Development of a System for Post-marketing Population Pharmacokinetic and Pharmacodynamic Studies using Real-World Data from Electronic Health Records. Clinical Pharmacology & Therapeutics. 2020 Apr;107(4):934-43. doi: 10.1002/cpt.1787.

