---
title: "Pro-Demographic"
description: |
  This tutorial describes how to process demographic data using *Pro-Demographic* module in the system.
author:
  - name: Michael L. Williams
output:
  distill::distill_article:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(R.options = list(width = 130))
library(EHR)
```

* See also "Pro-Demographic" in "[2. EHR Vignette for Structured Data](https://cran.r-project.org/web/packages/EHR/vignettes/ehr_vignette_02_str.html)" of `EHR` package.

# Introduction

This tutorial describes how to use the *Pro-Demographic* module to process demographic data (see Choi *et al.*$^{1}$ for details).

To begin we load the `EHR` package and create some directories to work from. 

```{r load-lib-dir, eval=FALSE}
# load EHR package
library(EHR)

td <- tempdir()
checkDir <- file.path(td, 'checks') # directory for interactive checking
dir.create(checkDir)
dataDir <- file.path(td, 'data') # directory for processed data
dir.create(dataDir)
rawDataDir <- system.file("examples", "str_ex2", package="EHR")
drugname <- 'fent'
LLOQ <- 0.05
```

# Demographic Data

We will use example demographic data to demonstrate the *Pro-Demographic* module. The raw data is shown below.

```{r}
demo.in <- read.csv(system.file("examples", "str_ex2","Demographics_DATA.csv", package="EHR"))[,c(1,2,3,4,5)]
head(demo.in,10)
```

Each row in the data represents an individual surgical date.  In this case, each individual is assigned a unique ID and has a recorded gender, weight, and height for each surgery.  The goal of *Pro-Demographic* is to make the demographic data suitable for merging with medication data for population pharmacokinetic analysis while apply exclusion criteria in a reproducible manner which preserves and organizes the original data.

# Preparing the Demographic Data

```{r}
td <- tempdir()
dir.create(file.path(td, 'data2'))
dataDir <- file.path(td, 'data2')
saveRDS(demo.in, file=file.path(dataDir,"demo.rds"))
```

# Running `run_Demo`

`run_Demo` will configure the demographic data, apply exclusion criteria and saving a record of exclusion rules/excluded individuals.  The argument `toexclude` takes an expression evaluating to `TRUE`/`FALSE` indicating which individuals will be excluded from the final dataset.

```{r}
low_wgt <- function(x, val=1) {  x < 6 }
demo.out <- run_Demo(demo.path = file.path(dataDir, "demo.rds"),
                     toexclude = expression(low_wgt(weight)))
```

# Processed Data

The demographic data is now ready for further processing.

```{r}
head(demo.out)
```


# References
1. Choi L, Beck C, McNeer E, Weeks HL, Williams ML, James NT, Niu X, Abou-Khalil BW, Birdwell KA, Roden DM, Stein CM. Development of a System for Post-marketing Population Pharmacokinetic and Pharmacodynamic Studies using Real-World Data from Electronic Health Records. Clinical Pharmacology & Therapeutics. 2020 Apr;107(4):934-43. doi: 10.1002/cpt.1787.
